---
title: "p8105 Midterm Project"
author: "Clement Mugenzi"
date: "10/17/2019"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(viridis)
library(wordcountaddin)
library(readxl)
library(readr)
library(patchwork)
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```


# Introduction

This project uses data analysis concepts to expand on a study that have been conducted examining the association between smart device use and enlarged External Occipital Protuberance in millenials.  


# Data Description

```{r}
hornlike_df = 
  read_excel("Data/p8105_mtp_data.xlsx", 
             range = "A9:I1230") %>% 
  janitor::clean_names() %>% 
# changing to appropriate variable names.
  rename(eop_group = "eop_size", eop_size = "eop_size_mm",
         fhp_size = "fhp_size_mm", fhp_group = "fhp_category",
         enlarged_eop = "eop_visibility_classification") %>%
# replacing missing values in the eop_size variable by 0.
# Creating character and ordered factors for categorical variables.
  mutate(
    eop_size = replace_na(eop_size, 0),
    sex = recode(sex, "0" = "Female", "1" = "Male"),
    age_group = recode(age_group,"2" = "18-30", 
                       "3" = "31-40", "4" = "41-50",
                       "5" = "51-60", "6" = "60+",
                       "7" = "60+", "8" = "60+"),
    eop_group = recode(eop_group, "0" = "0-5", "1" = "5-10", 
                       "2" = "10-15", "3" = "15-20",
                        "4" = "20-25", "5" = "25+"),
    fhp_group = recode(fhp_group, "0" = "0-10", "1" = "10-20",
                       "2" = "20-30", "3" = "30-40", 
                       "4" = "40-50", "5" = "50-60",
                       "6" = "60+", "7" = "60+"),
    enlarged_eop = recode(enlarged_eop, "0" = "Absent",
                            "1" = "Visible", "2" = "Fully Developed"),
    sex = factor(sex, levels = c("Female", "Male")),
    enlarged_eop = factor(enlarged_eop, levels = c("Absent",
                                                       "Visible",
                                                       "Fully Developed")),
    eop_shape = factor(eop_shape, levels = c("0", "1", "2", "3")),
    age_group = factor(age_group, levels = c("18-30", "31-40",
                                              "41-50", "51-60",
                                              "60+")),
    eop_group = factor(eop_group, levels = c("0-5", "5-10",
                                             "10-15", "15-20",
                                             "20-25", "25+")),
    fhp_group = factor(fhp_group, levels = c("0-10", "10-20",
                                             "20-30", "30-40",
                                             "40-50", "50-60",
                                             "60+"))) %>%
  arrange(sex) %>% 
  select(-eop_shape)
```


This dataset is comprised of **`r ncol(hornlike_df)`** variables and **`r nrow(hornlike_df)`** observations which is also the number representative of how many participants were studied by the original scientific report. In brief, I started by renaming relevant variables with appropriate names then used **mutate** function to recode values and redefine **sex, eop_visibility, age_group, eop_group, and fhp_group** variables by making them factors. 

```{r}
# Age and Gender Distribution
age_gender = 
  hornlike_df %>% 
  group_by(sex, age_group) %>% 
# renaming age group here but this won't affect the main dataframe.
  rename("Age Group" = "age_group") %>% 
  drop_na() %>%  
  summarise(
    frequency = n()) %>% 
  pivot_wider(
    names_from = sex,
    values_from = frequency) %>% 
  knitr::kable()
age_gender
```


## Issues with the Dataset

```{r}
# EOP Size and Group
hornlike_df %>% 
# Dropping missing values for 0-18 age group.
  drop_na() %>% 
  ggplot(aes(x = eop_group, y = eop_size, color = eop_size)) +
  geom_point() +
  labs(
    title = "Figure 1. EOP Size Versus EOP Group",
    x = "EOP Group",
    y = "EOP Size",
    color = "EOP Size")
```

Here the categorical variable **EOP Group** is misclassifying several **EOP Size** values. For instance, in the range **20-25**, the eop size seems to span values ranging from 20 to about 35 mm.


```{r}
# FHP Size and Group
hornlike_df %>% 
# Dropping missing values for 0-18 age group.
  drop_na() %>% 
  ggplot(aes(x = fhp_group, y = fhp_size, color = fhp_size)) +
  geom_point() +
  labs(
    title = "Figure 2. FHP Size Versus FHP Group",
    x = "FHP Group",
    y = "FHP Size",
    color = "FHP Size")
```

The same pattern of misclassification is happening here as well where FHP sizes are being classified in groups they should not be classified in due to poor definitions of ranges on the part of the researcher. For instance there is a FHP size belonging to the **60+** age group now being classified in the **40-50** age group.


# Visualization

```{r}
# Fig.3 and Fig.4
fig_3 = 
  hornlike_df %>% 
  drop_na() %>% 
  ggplot(aes(x = fhp_size, fill = age_group)) +
  geom_histogram(position = "dodge", binwidth = 3) +
  guides(fill = guide_legend(title = "Age Group")) +
  labs(
    title = "Distribution of FHP Size with respect to Sex",
    x = "FHP Size",
    y = "Count",
    caption = "Figure.3 Size between 0 and 89.3")

fig_4 = 
  hornlike_df %>% 
  group_by(enlarged_eop, age_group, sex) %>% 
  filter(enlarged_eop == "Fully Developed") %>% 
  summarise(
    num_eeop = n()) %>% 
  mutate(
    percent_eeop = num_eeop/sum(num_eeop) * 100) %>% 
  filter(num_eeop != 1) %>% 
  ggplot(aes(x = age_group, y = percent_eeop, color = sex, group = sex)) +
  geom_point() + 
  geom_line() +
  labs(
    title = "Figure.4 Rate of EEOP per Age Group",
    x = "Age Group",
    y = "Rate (Percent)",
    color = "Sex"
  )

fig_3 / fig_4
```

From **Figure 3**, we can clearly conclude that most FHP sizes are concentrated in the **18-30** age group and within **10-30** mm FHP range. And from **Figure 4**, males have an overall high percentage of EEOP compared to females.

```{r}
# 2x5 Collection of Panels
male_panel = 
  hornlike_df %>% 
  drop_na() %>% 
  filter(sex == "Male") %>% 
  ggplot(aes(x = fhp_size, y = eop_size, color = age_group)) +
  facet_grid(~age_group) + 
  geom_point() + 
  theme(legend.position = "none") +
  geom_smooth(color = "black", linetype = 1) +
  labs(
    title = "Association between FHP and EOP sizes in Men",
    x = "FHP Size (mm)",
    y = "EOP Size (mm)"
  )


female_panel = 
  hornlike_df %>% 
  drop_na() %>% 
  filter(sex == "Female") %>% 
  ggplot(aes(x = fhp_size, y = eop_size, color = age_group)) +
  facet_grid(~age_group) +
  geom_point() +
  geom_smooth(color = "black", linetype = 1) +
  labs(
    title = "Association between FHP and EOP sizes in Women",
    x = "FHP Size (mm)",
    y = "EOP Size (mm)",
    color = "Age Group"
  )

male_panel / female_panel 
```

There seems to be a positive association between EOP and FHP sizes for female subjects in the 18-30 age group but a steady pattern with few bumps in the remaining age groups. For males, there is an overall increasing pattern of EOP as FHP increases. 


# Reproducing Results

The research's sample size of **1200** is inconsistent with my sample size of 
**`r nrow(hornlike_df)`.** 

```{r}
# Mean and Standard Deviation for male subjects.
Male_stats = 
  hornlike_df %>% 
  filter(sex == "Male") %>% 
  summarise(
    Mean = mean(fhp_size, na.rm = T), 
    SD = sd(fhp_size, na.rm = T)) %>% 
  knitr::kable(digits = 1)
Male_stats
```



```{r}
# Mean and Standard Deviation for female subjects.
Female_stats = 
  hornlike_df %>% 
  filter(sex == "Female") %>% 
  summarise(
    Mean = mean(fhp_size, na.rm = T), 
    SD = sd(fhp_size, na.rm = T)) %>% 
  knitr::kable(digits = 1)
Female_stats
```

The reported mean and standard deviation for Male and Female sex were **28(15) mm** and **24(11) mm** respectively. This is consistent with my results. 


A subject is said to have **Enlarged Occipital Protuberance (EEOP)** only when the EOP exceeds **10 mm** in size, where EOP size is the key variable to determine EEOP prevalence.

```{r}
# Determining EEOP Prevalence
hornlike_df %>% 
  mutate(
    Condition = case_when(
      eop_size >= 10 ~ "Positive",
      eop_size < 10  ~ "Negative",
      TRUE           ~ "")) %>%
  group_by(Condition) %>%
  filter(Condition == "Positive") %>%
  summarise(
    EEOP = n(),
    Prevalence = (EEOP / 1221) * 100) %>%
  knitr::kable(digits = 1)
```

Therefore, the research's **33%** prevalence is quite consistent with my **32%** prevalence.

```{r}
# FHP size in older subjects
hornlike_df %>% 
  mutate(
    Common = case_when(
      fhp_size >  40 ~ "> 40 mm",
      fhp_size <= 40 ~ "< 40 mm",
      TRUE           ~ "")) %>%
  group_by(Common, age_group) %>%
  filter(Common == "> 40 mm") %>%
  drop_na() %>% 
  summarise(
    Frequency = n()) %>%
  mutate(
    Percentage = Frequency / sum(Frequency) * 100) %>%
  rename("Age Group" = age_group) %>% 
  knitr::kable(digits = 1)
```

The **57.6%** of FHP size  **> 40 mm** in 60 and older subjects is not constistent with this study's **34.5%.**

# Discussion

In conclusion, my results confirm a high EEOP rate in male subjects across all age groups and no significant association between EEOP and FHP size. Overall, the quality of the original report was poor because it finds higher FHP values are found in 60+ age group then concludes that a high prevalence of EEOP is found in the 18-30 age group. We cannot confirm that the use of smart devices is associated with EEOP/horn growth since there is no predictor detailing the time spent on smart devices. Therefore, time spent on smart devices will further help us reach a more telling conclusion.  





































